---
title: "What if IPC PIN Estimates Captured Their Temporal Evolution Instead of Offering a Static Snapshot?"
subtitle: "How This Could Help Nutrition Actors Respond More Effectively"
author: "Tomás Zaba"
date: today
format: revealjs
---

```{r}
#| label: source-script
Sys.setenv(path_secret_key = "~/.ssh/id_rsa")
source(file = "script.R")
```

## Context

<br> 

+ To prevent that a child dies from AMN, anticipdated information on the evolution of AMN is essential

+ Current approaches - (NxPxK) and now programme-based - do not provide such information

    - Not even possible to provide # cases by month, quarters, etc.

::: {.callout-note}

Recently, IPC users and decision-makers criticized the fact that IPC estimates were not sensitive to the changes forecasted in the projection period.

:::

# What if all this could be addressed, eventually? 🧐

# Insights from a time series forecasting model using Somalia's six years of admissions data {.smaller}

## Pastoral Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-pasto
forecast_pasto_plot
```

## Pastoral Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-pasto-jan-dec
forecast_pasto |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Agro-pastoral Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-agropasto
forecast_agropasto_plot
```

## Agro-pastoral Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-agropasto-jan-dec
forecast_agropasto |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Riverine Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-riverine
forecast_riverine_plot
```

## Riverine Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-riverine-jan-dec
forecast_riverine |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Urban/IDPs Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-urban/idps
forecast_urbanidps_plot
```

## Urban/IDPs Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-urban-idps-jan-dec
forecast_urbanidps |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

# The above is just the caseload, not the actual burden or PIN... {.smaller}

<br> 

$PIN = \text{forecast} * \frac{\text{1}}{\text{programme coverage}}$



## Advantages and Disadvantages {.smaller}

:::: {.columns}

::: {.column width="50%" style="text-align: left"}
### Advantages
+ It uses programme-based-only data to estimate caseloads
+ It provides estimates by month
    - Makes possible to get **estimates for current** and for **projection** by taking the forecasts in the period that corresponds to current and projection, respectively
+ Useful for fundraising, and to scale-up programme response for when a rise is expected
+ Useful for risk monitoring
:::

::: {.column}
### Disadvantages
+ The model used does not account for incoming shocks, but there are other models to be explored
:::

:::: 

# How are you processing all of this? 🤨 🧐