---
title: "What if IPC PINs moved with the months?"
subtitle: "Helping nutrition actors respond more effectively to changing needs"
format: revealjs
---

```{r}
#| label: source-script
Sys.setenv(path_secret_key = "~/.ssh/id_rsa")
source(file = "script.R")
```

## Context {.smaller}

<br> 

+ To prevent that a child dies from AMN, actionable information on the evolution of AMN is essential to inform decision-making

+ Current approaches for estimating people (children) in need (of treatment due to acute malnutrition): 
  - Using the "NxPxK" formulae
  - Programme-based data, as in 

Do not provide such information ❌

::: {.callout-note}

Recently, following the April analysis for Gaza, Nutrition Cluster and Food Security partners — along with decision-makers more broadly — criticised the fact that IPC AMN estimates did not reflect the changes projected.

  + Despite an anticipated deterioration in the acute malnutrition situation over the coming months, the AMN PIN remained unchanged.

:::

# What if all this could be addressed? 🧐 {.smaller}

# Insights from a time series forecasting model using Somalia's six years of severe acute malnutrition admission data {.smaller}

## Pastoral Livelihood System (1/2)
### Time plot

```{r}
#| label: forecast-pasto
forecast_pasto_plot
```

## Pastoral Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-pasto-jan-dec
forecast_pasto |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Agro-pastoral Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-agropasto
forecast_agropasto_plot
```

## Agro-pastoral Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-agropasto-jan-dec
forecast_agropasto |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Riverine Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-riverine
forecast_riverine_plot
```

## Riverine Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-riverine-jan-dec
forecast_riverine |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

## Urban/IDPs Livelihood System (1/2)
### Time plot
```{r}
#| label: forecast-urban/idps
forecast_urbanidps_plot
```

## Urban/IDPs Livelihood System (2/2)
### A glance at January to December 2025

```{r}
#| label: forecast-urban-idps-jan-dec
forecast_urbanidps |>
    filter(year(Monthly) == 2025) |> 
    ggplot() +
    geom_ribbon(
        aes(x = Monthly, ymin = lower, ymax = upper, fill = level),
        alpha = 0.5
        ) +
    geom_line(
        aes(x = Monthly, y = mean_inv, colour = "Forecast mean"),
        alpha = 0.6
        ) +
  scale_fill_manual(
    name = "Confidence Interval",
    values = c("80%" = "#1F77B4", "95%" = "#AEC7E8")
  ) +
  scale_colour_manual(
    name = "Series",
    values = c(
      "Forecast mean" = "#0072B2"
    )
  ) +
  labs(
    title = "Forecasted SAM admissions into the programme",
    subtitle = "Time horizon: from January to December 2025",
    y = "Number of cases",
    x = "Monthly[1M]"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 10),
    plot.subtitle = element_text(size = 9, colour = "#706E6D"),
    axis.title.y = element_text(size = 10, margin = margin(r = 5)),
    axis.title.x = element_text(size = 10, margin = margin(r = 5))
  )
```

# What could IPC AMN look like if it embraced this approach? 🧐 {.smaller}

## {.smaller}

<br> 

+ The forecast model provides the following ouputs: 
  - **A forecast mean value**: the point forecast
  - **Lower and Upper value of the confidence intervals** 

+ These are not the actual PIN

 - It can be gotten by applying the formulae: 

$PIN = \text{forecast} * \frac{\text{1}}{\text{programme coverage}}$

<br> 

+ **where**:  

  - *forecast* is the point forecast  
  - *programme coverage* is the Nutriton Cluster-based value (%) used to ascertain the programme reach based on available resources

+ This would yield into 👉 

## A glance at the forecasted values {.smaller}

```{r}
#| label: how-to-get-pin
x <- forecast_pasto |> 
  select(lsystems, month = Monthly, mean_inv, level, lower, upper) |> 
  subset(level == "80%") |> 
  mutate(
    forecasts = round(mean_inv, 0),
    lower = round(lower),
    upper = round(upper),
    PIN = forecasts * (1/0.5) |> round(),
    PIN_lower = lower * (1/0.5) |> round(),
    PIN_upper = upper * (1/0.5) |> round()
  ) |> 
    relocate(forecasts, .before = lower) |> 
    select(-3)

x |> knitr::kable()
```

## {.smaller}
### Then, the PIN can be split by IPC validity periods... {.smaller}

+ Assuming the parameters in a given analysis: **Current: January to May**; **Projection: June to December**

:::: {.columns}

::: {.column width="50%" style="text-align: left"}

**Current: January to May 2025**
```{r}
#| label: pin-ipc-current
current <- x |> 
  subset(month < yearmonth("2025 June")) |> 
  select(month, PIN, PIN_lower, PIN_upper)

current |> 
  knitr::kable()
```

+ **Total PIN in Current: `r format(sum(current$PIN), big.mark = ",", scientific = FALSE)`**
:::

::: {.column style="text-align: right"}
**Projection: June to December 2025**
```{r}
#| label: pin-ipc-projection
proje <- x |> 
  subset(month >= yearmonth("2025 June")) |> 
  select(month, PIN, PIN_lower, PIN_upper)

proje |> 
  knitr::kable()
```

+ **Total PIN in Projection: `r format(sum(proje$PIN), big.mark = ",", scientific = FALSE)`**
:::

::::

<br>


## {.smaller}
### Accouting for changes in the projection analysis...

* With the point forecasts and upper and lower levels of the confidence intervals...

  + In areas in which: 
    - a **deterioration** is expected: the upper limit of the forecast could be used
    - an **improvement** is expected: the lower limit of the forecast could be used
    - **neither of the above is expected** (similar): the point forecast could be used

::: {.callout}
This would help balancing the PIN, and eventually a breakthrough to what we do now, and a fix the current, and outstanding, issue: one number for the whole year.

  + Possibly, a fix to revising the PIN in projection updates
  + Useful for fundraising, and to scale-up programme response for when a rise is expected
  + Useful for risk monitoring

:::


# Is this forecasting approach user-friendly enough?

 + NO! It is technically demanding...

# Could it be automated — or at least semi-automated? 

+ ABSOLUTELY!

# How are you processing all of this? 🤨 🧐